Financialisation of Commodity Markets

### 2024-09-25 docs/paper.qmd
Complete first draft of results for the post-crisis period

@tbl-stats-no-regimes displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five--Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)--show a lower figure with five significant figures--Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)--against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  
The pattern of results for volatilities is similar although the increase is less pronounced. Both equally weighted portfolios shows an increase, more pronounced for the GB assets, 10% to 13.6% (US) vs. 15.7% to 26.9% (GB), while out of the thirty assets, eleven show a drop in volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C (IFUS), Lumber (XCME), Sugar-#11 (IFUS), Natural gas (XNYM), Crude oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM). Financialisation seems however to have impacted the commodity complex beyond these mechanical effects of increased long participation ensuing from the entry on index type investments as our results suggest next.

The cross-sectional nature of financialisation naturally leads us to extend the individual hedging pressure paradigm by incorporating aggregate CHP (**CHP**), a market wide measure of hedging pressure. The traditional Keynesian hedging pressure paradigm postulates a negative relationship between the return on an individual commodity futures and its hedging pressure. The extended Keynesian hedging pressure paradigm, in the context of financialisation, should also imply that returns for individual commodities should be high in periods where **CHP** is low and vice et versa. In @tbl-stats-regimes the mean returns are shown over high (**contango**) and low (**backwardation**) **CHP** regimes for each time period.  
Over the first period eighteen of the twenty-four US commodities have higher mean returns during periods of **backwardation** as compared to **contango**, with the difference being statistically significant at the 10% level for two commodities, at the 5% level for one and at the 1% level for another. For the remaining six commodities the higher mean return during **contango** is not statistically significant for any of them. For the six GB commodities, all metals, five of them had higher mean returns during phases of **backwardation** as compared to **contango** in the first period, with one of these differences being statistically significant at the 10% level while for the remaining asset (Zinc (XLME)) the higher mean return during **contango** is not statistically significant. An equally weighted portfolio of the twenty-four US commodities had a mean return of 15.4% during **backwardation** phases, statistically significant at the 1% level, and a mean return of -1.1% during **contango**, both in the first period, with the difference between them statistically significant at the 5% level. The corresponding figures for an equally weighted portfolio of the GB metals were 4.4% and 0.3% respectively. These results provide broad support for an aggregate version of the Keynesian hedging pressure hypothesis and suggest that hedgers in the aggregate were engaged in risk transfer during this period.  
Over the second period the pattern essentially reverses. For the twenty-four US commodities over this period, fourteen have higher mean returns during phases of **contango** over **backwardation**. Four of the six GB metals achieved higher mean returns over **contango** relative to **backwardation** phases with the difference being statistically significant for one commodity. The equally weighted US portfolio had a mean return of 15.7% over **backwardation** against 16.2% during phases of **contango**, both statistically significant at the 10% level. The difference was considerably more pronounced for the equally weighted GB metals portfolio with a mean return of 8% during **backwardation** rising to 35.6% during **contango**.

@@

The entire cross-section sees an increase in average regression coefficient relative to the market factor over the crisis period with the strongest progression for the GB traded metals followed by the agricultural sector (@tbl-regressions-index). The pricing performance of the factor 
dominates the US commodity complex over the period, showing a 50% increase in mean average $R^{2}$ to over 31%, while that for both the CHP and open interest factor decreases and, although increasing, remains modest for the term structure factors; the increase in strongest for the GB traded metal assets with an average $R^{2}$ of almost 34% for the period. Taken together, these results indicate the presence of a systematic factor across the entire cross-section of the US commodities. For both the CHP and term structure factors, both legs show an increase in average $R^{2}$ and while substantially diverging over the preceding financialisation period the pricing performance for both legs seem to converge back to the pattern observed pre-financialisation.

@@

Taken together the patterns of results over the crisis and post-crisis periods echo those in @basu_definancialisation in suggesting that the exogenous shock delivered by the financial crisis combined with the accommodative monetary policy that ensued seem to have initiated a reversion toward pre-financialisation fundamentals in the commodity futures complex.

@@

```{r}
#| label: tbl-stats-no-regimes
#| tbl-cap: "This table shows mean returns and volatility (sd) for the twenty four individual US commodities and the six GB metals considered in the study as well as for two equally weighted portfolios formed from the US commodities and the GB metals respectively across the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). Mean values significant at the 1%, 5% and 10% level are marked with \\\\*\\\\*\\\\*, \\\\*\\\\* and \\\\* respectively. The results are discussed in section \\@sec-results."

@@

```{r}
#| label: tbl-stats-regimes
#| tbl-cap: "This table shows mean returns and volatility (sd) for the twenty four individual US commodities and the six GB metals considered in the study as well as for two equally weighted portfolios formed from the US commodities and the GB metals respectively across the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). Figures are shown independently for phases of aggregate backwardation (aggregate CHP $\\leq$ period median) and aggregate contango (aggregate CHP > period median). Mean values significant at the 1%, 5% and 10% level are marked with \\\\*\\\\*\\\\*, \\\\*\\\\* and \\\\* respectively. Aggregate CHP construction and corresponding regime definitions are discussed in section \\@sec-methods while the results are discussed in section \\@sec-results."

@@

```{r}
#| label: tbl-time-reg-US-CHP
#| tbl-cap: "This table shows the average time series $R^{2}$ for models where the returns series for the twenty four individual US commodities are independently regressed against relative change in their own commercial hedging pressure (CHP) series as well as against that in aggregate CHP, a market wide measure of CHP which calculation method is discussed in section \\@sec-methods. Averages are calculated independently across the whole cross-section of US traded commodity assets considered in the study as well as across commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). The results are discussed in section \\@sec-results."

@@

```{r}
#| label: tbl-correlations-periods
#| tbl-cap: "This table shows the average pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as across countries and commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results are calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Aggregate CHP construction and corresponding regime definitions are discussed in section \\@sec-methods while the results are discussed in section \\@sec-results."

@@

```{r}
#| label: tbl-correlations-years
#| tbl-cap: "This table shows the average pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as across countries and commodity sectors and subsectors by year for the whole period of interest (1997-2018). The results are discussed in section \\@sec-results."

@@

```{r}
#| label: tbl-regressions-index
#| tbl-cap: "This table shows the average time series regression coefficients for models where the returns series for all the commodity assets considered in the study are independently regressed against the return series for our market factor, an equally weighted portfolio of the 24 US commodities considered in the study. The results are calculated independently across countries as well as commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). The market factor construction is discussed in section \\@sec-methods while the results are discussed in section \\@sec-results."

@@

```{r}
#| label: tbl-regressions-factors
#| tbl-cap: "This table shows the average time series regression $R^{2}$ for models where the returns series for all the commodity assets considered in the study are independently regressed against the return series for our mimicking portfolios for risk factors in returns, including the market, CHP, open interest and term structure factors. The results are calculated independently across countries and commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018); for each period the results are calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Factors as well as aggregate CHP construction and corresponding regime definitions are discussed in section \\@sec-methods while the results are discussed in section \\@sec-results."

>>>>

@tbl-stats-no-regimes displays descriptive statistics for the commodity series considered in the study; the results show a clear pattern of increase in average return during financialisation with an equally weighted portfolio of the US traded assets showing a mean returns of 15.9% significant at the 5% level vs. 6.9% with a 10% significance in the pre-financialisation period. The increase is even more pronounced for an equally weighted portfolio of GB traded assets (metals), 2.47% (not significant) to 21.8% (significant at the 10% level), which sets out the global nature of the phenomenon. Most commodities show a higher mean return while five--Lumber (XCME), Natural gas (XNYM), Palladium (XNYM), Platinum (XNYM) and Nickel-primary (XLME)--show a lower figure with five significant figures--Crude oil-WTI (XNYM), Heating oil (XNYM), Copper (XCEC), Copper (XLME) and Tin-refined (XLME)--against none pre-financialisation. Sugar-#11 wins the race with a mean return increasing from -4.6% to 24.9% followed by copper with sizeable increases for both markets: 1.5% to 28.2% (XCEC) and 0.5% to 28.6% (XLME).  
The pattern of results for volatilities is similar although the increase is less pronounced. Both equally weighted portfolios shows an increase, more pronounced for the GB assets, 10% to 13.6% (US) vs. 15.7% to 26.9% (GB), while out of the thirty assets, eleven show a drop in volatility: Oats (XCBT), Lean hogs (XCME), Cocoa (IFUS), Coffee-C (IFUS), Lumber (XCME), Sugar-#11 (IFUS), Natural gas (XNYM), Crude oil-WTI (XNYM), Gasoline (XNYM), Heating oil (XNYM), Palladium (XNYM). Financialisation seems however to have impacted the commodity complex beyond these mechanical effects of increased long participation ensuing from the entry on index type investments as our results suggest next.   
The cross-sectional nature of financialisation naturally leads us to extend the individual hedging pressure paradigm by incorporating aggregate CHP (**CHP**), a market wide measure of hedging pressure. The traditional Keynesian hedging pressure paradigm postulates a negative relationship between the return on an individual commodity futures and its hedging pressure. The extended Keynesian hedging pressure paradigm, in the context of financialisation, should also imply that returns for individual commodities should be high in periods where **CHP** is low and vice et versa. In @tbl-stats-regimes the mean returns are shown over high (**contango**) and low (**backwardation**) **CHP** regimes for each time period.  
Over the first period eighteen of the twenty-four US commodities have higher mean returns during periods of **backwardation** as compared to **contango**, with the difference being statistically significant at the 10% level for two commodities, at the 5% level for one and at the 1% level for another. For the remaining six commodities the higher mean return during **contango** is not statistically significant for any of them. For the six GB commodities, all metals, five of them had higher mean returns during phases of **backwardation** as compared to **contango** in the first period, with one of these differences being statistically significant at the 10% level while for the remaining asset (Zinc (XLME)) the higher mean return during **contango** is not statistically significant. An equally weighted portfolio of the twenty-four US commodities had a mean return of 15.4% during **backwardation** phases, statistically significant at the 1% level, and a mean return of -1.1% during **contango**, both in the first period, with the difference between them statistically significant at the 5% level. The corresponding figures for an equally weighted portfolio of the GB metals were 4.4% and 0.3% respectively. These results provide broad support for an aggregate version of the Keynesian hedging pressure hypothesis and suggest that hedgers in the aggregate were engaged in risk transfer during this period.  
Over the second period the pattern essentially reverses. For the twenty-four US commodities over this period, fourteen have higher mean returns during phases of **contango** over **backwardation**. Four of the six GB metals achieved higher mean returns over **contango** relative to **backwardation** phases with the difference being statistically significant for one commodity. The equally weighted US portfolio had a mean return of 15.7% over **backwardation** against 16.2% during phases of **contango**, both statistically significant at the 10% level. The difference was considerably more pronounced for the equally weighted GB metals portfolio with a mean return of 8% during **backwardation** rising to 35.6% during **contango**.

@@

The entire cross-section sees an increase in average regression coefficient relative to the market factor over the crisis period with the strongest progression for the GB traded metals followed by the agricultural sector (@tbl-regressions-index). The pricing performance of the factor 
dominates the US commodity complex over the period (@tbl-regressions-factors), showing a 50% increase in mean average $R^{2}$ to over 31%, while that for both the CHP and open interest factor decreases and, although increasing, remains modest for the term structure factors; the increase is strongest for the GB traded metal assets with an average $R^{2}$ of almost 34% for the period. Taken together, these results indicate the presence of a systematic factor across the entire cross-section of the US commodities. For both the CHP and term structure factors, both legs show an increase in average $R^{2}$ and while substantially diverging over the preceding financialisation period the pricing performance for both legs seems to converge back to a pattern closer to that observed pre-financialisation.

@@

The average pairwise correlations fall across the board in the post-crisis period, below the levels observed during financialisation, with the figure for the entire cross-section (@tbl-correlations-periods) dropping down to 0.12 from 0.28 over the crisis. This substantial decline, that does not appear to have been widely noted, appears to begin in the latter part of the crisis period, around 2012 (@tbl-correlations-years), before levelling down over the whole post-crisis period.

In a similar vein, the average market factor beta for the entire cross-section drops down to 0.94 (@tbl-regressions-index), below the level observed during financialization. The decline seems driven by the metals sector, particularly the GB traded segment, with the figures for the US and GB traded assets at 0.91 and 0.72, down from 1.12 and 1.18, respectively. Meanwhile, the pricing performance of the market factor also retreats to below financialisation levels with the average $R^{2}$ for the entire cross-section dropping down to 14.7% (@tbl-regressions-factors). All sectors seem affected with agriculturals, energy and metals all seeing their average $R^{2}$ decline below the levels observed during the financialisation period albeit remaining above 10% for the GB traded metals against less than 2% pre-financialisation. Similarly, the performance for both the CHP and term structure factors falls back to levels below those attained over financialisation while the gap between their long and short legs remains narrower than observed during that period. Taken together, these results, along with those from the crisis period, suggest that the exogenous shock of the financial crisis, combined with the subsequent accommodative monetary policy, may have triggered a reversion toward pre-financialization fundamentals in the commodity futures complex; an issue that, to our knowledge, has not been widely noted in the existing literature.  
Meanwhile the regime based analysis is also revealing starting with the average correlations are higher during phases of **contango** and the pattern stronger for the metals assets both US and GB traded. Similarly, for these the pricing performance of the market factor is considerably better during phase of **contango** in contrast with the previous periods that showed no substantial difference between phase of **contango** and **backwardation**. The figures for the US and GB traded segments of the sector are 23% (**contango**) vs 11.5% (**backwardation**) against 36.2% vs 43% during financialisation and 14.4% (**contango**) vs 6.6% (**backwardation**) against 33.5% vs 34.4% during financialisation respectively which indicate some recent changes in the correlation structure within the sector. Taken together with the decline in interdependence in the commodity sector at large documented above, this observation could form a valuable basis for future research.

@@

```{r}
#| label: tbl-stats-no-regimes
#| tbl-cap: "This table shows mean returns and volatility (sd) for the twenty four individual US commodities and the six GB metals considered in the study as well as for two equally weighted portfolios formed from the US commodities and the GB metals respectively across the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). Mean values significant at the 1%, 5% and 10% level are marked with \\\\*\\\\*\\\\*, \\\\*\\\\* and \\\\* respectively. The results are discussed in \\@sec-results."

@@

```{r}
#| label: tbl-stats-regimes
#| tbl-cap: "This table shows mean returns and volatility (sd) for the twenty four individual US commodities and the six GB metals considered in the study as well as for two equally weighted portfolios formed from the US commodities and the GB metals respectively across the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). Figures are shown independently for phases of aggregate backwardation (aggregate CHP $\\leq$ period median) and aggregate contango (aggregate CHP > period median). Mean values significant at the 1%, 5% and 10% level are marked with \\\\*\\\\*\\\\*, \\\\*\\\\* and \\\\* respectively. Aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results."

@@

```{r}
#| label: tbl-time-reg-US-CHP
#| tbl-cap: "This table shows the average time series $R^{2}$ for models where the returns series for the twenty four individual US commodities are independently regressed against relative change in their own commercial hedging pressure (CHP) series as well as against that in aggregate CHP, a market wide measure of CHP which calculation method is discussed in \\@sec-methods. Averages are calculated independently across the whole cross-section of US traded commodity assets considered in the study as well as across commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). The results are discussed in \\@sec-results."

@@

```{r}
#| label: tbl-correlations-periods
#| tbl-cap: "This table shows the average pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as across countries and commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). For each period, the results are calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results."

@@

```{r}
#| label: tbl-correlations-years
#| tbl-cap: "This table shows the average pairwise correlation coefficients calculated independently across the whole cross-section of commodity assets considered in the study as well as across countries and commodity sectors and subsectors by year for the whole period of interest (1997-2018). The results are discussed in \\@sec-results."

@@

```{r}
#| label: tbl-regressions-index
#| tbl-cap: "This table shows the average time series regression coefficients for models where the returns series for all the commodity assets considered in the study are independently regressed against the return series for our market factor, an equally weighted portfolio of the 24 US commodities considered in the study. The results are calculated independently across countries as well as commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018). The market factor construction is discussed in \\@sec-methods while the results are discussed in \\@sec-results."

@@

```{r}
#| label: tbl-regressions-factors
#| tbl-cap: "This table shows the average time series regression $R^{2}$ for models where the returns series for all the commodity assets considered in the study are independently regressed against the return series for our mimicking portfolios for risk factors in returns, including the market, CHP, open interest and term structure factors. The results are calculated independently across countries and commodity sectors and subsectors for the four periods of interest (past: 1997-2003; financialisation: 2004-2008; crisis: 2008-2013; post-crisis: 2013-2018); for each period the results are calculated independently over the entire period as well as over phases of aggregate backwardation and contango. Factors as well as aggregate CHP construction and corresponding regime definitions are discussed in \\@sec-methods while the results are discussed in \\@sec-results."
